Dissertation Proposal

### 2023-11-07 proposal.tex
rq2

To find a reasonably comprehensive collection of misconceptions, I
seek for inputs from students. Most prior works
(\Cref{s:background-misconceptions}) have experts coming up with a
list of potential misconceptions and then filter the potential
misconceptions with student inputs. This approach can't find
misconceptions not envisioned by experts and hence is hindered by
experts' blind spots.

@@

\subsection{Expected Results}

Eventually, I expect to build a table of misconceptions. Each
misconception comes with multiple programs. For each of these
programs, the wrong result that correspond to the misconception should
be commonly predicted by students from various population.
Furthermore, these students should give similar explanations for all
the wrong answers. For example, I might find a misconception called
\miscon{DefByRef}: when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; and when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}). My
current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful.

I have created a tutor (\Cref{s:system-tutor}) that incorporates all
these ideas (\Cref{s:tutor-uiux}) and have collected data from all
populations. I plan to explore more literature and to improve the
tutor if I find new helpful techniques.

@@

@@

Let's reconsider this program from \Cref{s:plan-rq1}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which correspond
to the \miscon{DefByRef} misconception. If an MCQ only has a small
number of options like this example, students have a great chance (in
this case, 50\%) of just guessing the correct answer or successfully
using a process of elimination. So I have been adding additional
options that are not obviously wrong. The tutor is adding options with
an ad hoc rules. Eventually, I plan to generate options by moving
and/or copying numbers in the current options. In this case, my plan
would add \lstinline|12 12| and \lstinline|0 12|. In summary, each MCQ
has the following options:
\begin{itemize}
  \item the correct answer,
  \item wrong answers that correspond to known misconceptions,
  \item options generated by moving and/or copying numbers in the
        current options
  \item the ``Other'' option (which, once choosen, allows students to
        enter arbitrary answer)
\end{itemize}

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor has been giving panelty. Designing
the strategy of assigning panelty is non-trivial. A common strategy is
to grade students by their correctness. However, grading by
correctness is not suitable for a tutoring system because we do not
mind students doing badly on earlier tasks as long as they eventually
master the concepts. Actually, if students generally succeed in all
tasks, the tutoring system is failing -- it does not teach anything to
the students. A alternative common strategy is to grade by completion.
However, grading by completion alone encourage students to rush
through the tutorials and hence might encourage students to make
random choices. In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, I graded by
completion\footnote{I have no control over how other institutions use
  the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

The tutor provides considerable opportunities for making case
comparison: each interpreting task presents a SMoL program and the
program's result. These program-result pairs are concrete cases of the
semantics of SMoL. To incorporate the idea of case comparsion, the
tutor currently prompts students to compare these cases and then to
summarize the rules of program behavior at the end of \emph{some}
tutorials. I plan to expand this to \emph{all} tutorials and use a
consistent prompt. The prompt should asks about the behavior of new
language constructs and their interaction with all previously
introduced constructs. For example, at the end of \textbf{mut-vars}, I
would want the tutor to ask
\begin{quote}
  What does \lstinline|set!| do? How does it interact with variables,
  \lstinline|defvar|, and \lstinline|deffun|?
\end{quote}

\subsection{Multi-syntax}%
\label{s:multi-syntax}

[FILL] Use a random default syntax per task? Log syntax switch.

@@

Prior research argues that illustrating how programs runs with
notional machines (NMs) is benefitial for students [CITE]. I have created a
web-based notional machine
\href{https://lukuangchen.github.io/stacker-2023}{https://lukuangchen.github.io/stacker-2023}.

@@

To fully answer my research questions, I should also try to prove that
my tutor is effective. In this secton, I describe how I want to refine
the design of the tutor to help me make the prove, what data I want to
collect, and what (statistical) analysis do I want to perform on the
data.

I would like to evaluate the effectiveness from two perspectives:
\begin{itemize}
  \item How effective is the tutor at stopping students from holding
        misconceptions?
\end{itemize}

@@

Given this observation, I plan to set up the tutor such that
\begin{itemize}
  \item Every wrong answer corresponds to up to one misconception. In
        this case, I say the wrong answer \emph{represents} the
        misconception.
  \item Every misconception is represented by some questions.
  \item In each tutorial, a misconception is either not represented by
        any interpreting tasks or represented by at least two tasks.
  \item Within each tutorial, the order of interpreting tasks is
        randomized.
\end{itemize}
For each tutorial and each misconception, I plan to do a model
comparison (with the standard AIC and BIC criteria) on the following
models/theories:
\begin{description}
  \item[TaskSpecific] Some students are more likely to choose the
  representing wrong answers in certain tasks.
  \item[PositionSpecific] Students are more likely to choose the
  representing wrong answer in a task if the task appears earlier (or
  later) in the tutorial.
\end{description}

If the \textbf{TaskSpecific} model best explains the data, I deduce
that some programs are more confusing and that we need further
investigation into the relationship between the tasks and misconceptions.

If the \textbf{PositionSpecific} model best explains \emph{and} students
perform better in later tasks, I deduce that the tutor seeems to be
correcting misconceptions. If studnets perform worse in later taks, the
tutor is potentially doing harm.

To know students well, I would like to have as many tasks per
misconception as possible. But in practice, the number of tasks is
subject to the following limitations:
\begin{enumerate}
  \item I don't want the length of any tutorial to exceed 30 minutes.
  Because the tutor is now asking for student explanation on choice,
  students are likely to spend more time on each task then before.
  \item Programs shouldn't be too long. Some misconceptions are really
  similar and as a consequence have only a small number of short
  programs to tell them apart.
\end{enumerate}

@@

In \Cref{s:background-misconceptions}, we discuss several papers that have provided
reports of student misconceptions with different fragments of
SMoL. However, it is difficult to know how comprehensive these
are. While some are unclear on the origin of their programs, they
generally seem to be expert-generated.

The problem with expert-generated lists is that they can be quite
incomplete. Education researchers have documented the
phenomenon of the \emph{expert blind spot} \citep{nathanExpertBlindSpot2001}: experts
simply do not conceive of many learner difficulties.
Thus, we need
methods to identify problems beyond what experts conceive.

>>>>

In \Cref{s:background-misconceptions}, I discuss several papers that
have provided reports of student misconceptions with different
fragments of SMoL. However, it is difficult to know how comprehensive
these are. While some are unclear on the origin of their programs,
they generally seem to be expert-generated.

The problem with expert-generated lists is that they can be quite
incomplete. Education researchers have documented the phenomenon of
the \emph{expert blind spot} \citep{nathanExpertBlindSpot2001}:
experts simply do not conceive of many learner difficulties.

@@

\subsection{Expected Results}%
\label{s:rq1-expected-results}

Eventually, I expect to build a table of misconceptions. Each
misconception comes with multiple programs. For each of these
programs, the wrong result that correspond to the misconception should
be commonly predicted by students from various population.
Furthermore, these students should give similar explanations for all
the wrong answers. For example, I might find a misconception called
\miscon{DefByRef} (See
\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
in \Cref{s:misconceptions-tutor} for a list of misconceptions that I
have found): when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; and when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}). My
current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a tutor
(\Cref{s:system-tutor}) that incorporates all these ideas
(\Cref{s:tutor-uiux}) and have collected data from all populations. I
plan to explore more literature and to improve the tutor if I find new
helpful techniques.

@@

Each tutorial might be delivered as multiple sections to avoid forcing
students to continuesly work for more than 30 minutes.

@@

Let's reconsider this program from \Cref{s:rq1-expected-results}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which correspond
to the \miscon{DefByRef} misconception. If an MCQ only has a small
number of options like this example, students have a great chance (in
this case, 50\%) of just guessing the correct answer or successfully
using a process of elimination. So I have been adding additional
options that are not obviously wrong. The tutor is adding options with
an ad hoc rules. Eventually, I plan to generate options by moving
and/or copying numbers in the current options. In this case, my plan
would add \lstinline|12 12| and \lstinline|0 12|. In summary, each MCQ
has the following options:
\begin{itemize}
  \item the correct answer,
  \item wrong answers that correspond to known misconceptions,
  \item options generated by moving and/or copying numbers in the
        current options
  \item the ``Other'' option (which, once choosen, allows students to
        enter arbitrary answer)
\end{itemize}

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor has been giving panelty. Designing
the strategy of assigning panelty is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because we do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item A alternative common strategy is to grade by completion.
  However, grading by completion alone encourage students to rush
  through the tutorials and hence might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

The tutor provides opportunities for making case comparison: each
interpreting task presents a SMoL program and the program's result.
These program-result pairs are concrete cases of the semantics of
SMoL. To incorporate the idea of case comparsion, the tutor currently
prompts students to compare these cases and then to summarize the
rules of program behavior at the end of \emph{some} tutorials. I plan
to expand this to \emph{all} tutorials and use a consistent prompt.
The prompt should asks about the behavior of new language constructs
and their interaction with all previously introduced constructs.

The current scope tutorial uses the following prompt
\begin{quote}
  When we see a variable reference, how do we find the corresponding
  declaration, if any?
\end{quote}
This prompt has a few problems:
\begin{itemize}
  \item It is unclear to adapt it to other tutorials.
  \item Students tend to give brief response possibly because the
  questions is asking for a lot and because students are tired when
  they reach the end of a tutorial.
\end{itemize}.

So I plan to change the question format into a notice-and-wonder style
of questions, which is known to be less demanding
(\Cref{s:pedagogic-techniques}). So I plan to replace the
\textbf{scope} prompt with
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|defvar|s work?
    \item How do \lstinline|deffun|s work?
    \item How do variable referencess (e.g., \lstinline|x| in \lstinline|(+ x 1)|) work?
    \item How do \lstinline|defvar|s, \lstinline|deffun|s, and variable references interact with each other?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{response with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{response with} what you are uncertain about these language constructs.
\end{quote}
I plan to use a similar prompt in later tutorials. For example, in \textbf{mut-vars},
I would want the tutor to ask
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|set!|s work?
    \item How do \lstinline|set!|s interact with other language constructs (namely, \lstinline|defvar|s, \lstinline|deffun|s, and variable references)?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{response with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{response with} what you are uncertain about these language constructs.
\end{quote}
I am hoping these new prompt are

@@

Prior research argues (\Cref{s:pedagogic-techniques})that illustrating
how programs runs with notional machines (NMs) is benefitial for
students. I have created a web-based notional machine
\begin{center}
  \href{https://lukuangchen.github.io/stacker-2023}{https://lukuangchen.github.io/stacker-2023}
\end{center}

@@

To fully answer my research questions, I should also argue that my
tutor is effective at correcting misconceptions. In this secton, I
describe how I want to refine the design of the tutor to help me make
the argument, what data I want to collect, and what (statistical)
analysis do I want to perform on the data.

@@

Given this observation, I plan to set up the tutor such that
\begin{itemize}
  \item Every wrong answer corresponds to up to one misconception. In
        this case, I say the wrong answer \emph{represents} the
        misconception.
  \item Every misconception is represented by some questions.
  \item In each tutorial, a misconception is either not represented by
        any interpreting tasks or represented by at least two tasks.
  \item Within each tutorial, the order of interpreting tasks is
        randomized.
\end{itemize}
For each tutorial and each misconception, I plan to do a model
comparison (with the standard AIC and BIC criteria) on the following
models/theories:
\begin{description}
  \item[TaskSpecific] Some students are more likely to choose the
  representing wrong answers in certain tasks.
  \item[PositionSpecific] Students are more likely to choose the
  representing wrong answer in task that appears earlier (or later) in
  the tutorial.
\end{description}

If the \textbf{TaskSpecific} model best explains the data, I deduce
that some programs are more confusing and that we need further
investigation into the relationship between the tasks and
misconceptions. If the \textbf{PositionSpecific} model best explains
\emph{and} students perform better in later tasks, I deduce that the
tutor seeems to be correcting misconceptions. If studnets perform
worse in later taks, the tutor is potentially doing harm.

To know students well, I would like to have as many tasks per
misconception as possible. But in practice, the number of tasks is
subject to the following limitations:
\begin{enumerate}
  \item I don't want the length of any tutorial section to exceed 30
  minutes.
  \item Programs shouldn't be too long. Some misconceptions are really
  similar and as a consequence have only a small number of short
  programs to tell them apart.
\end{enumerate}

@@
